Parallel Adaptive Importance Sampling

### 2018-02-28 PAIS.tex
Final edits and spellcheck complete

\author{Colin Cotter\thanks{Department of Mathematics, Imperial
    College, London, UK} \and Simon Cotter\thanks{School of
    Mathematics, University of Manchester, Manchester, UK. e:
    simon.cotter@manchester.ac.uk. SLC is grateful for EPSRC First
    grant award EP/L023393/1} \and Paul Russell\thanks{School of
    Mathematics, University of Manchester, Manchester, UK}}
\title{Parallel Adaptive Importance Sampling}
\begin{document}
\maketitle
\begin{abstract}
  Markov chain Monte Carlo methods are a powerful and commonly used
  family of numerical methods for sampling from complex probability
  distributions. As applications of these methods increase in size and
  complexity, the need for efficient methods which can exploit the
  parallel architectures which are prevalent in high performance
  computing increases. In this paper, we aim to develop a framework
  for scalable parallel sampling algorithms. At each iteration, an
  importance sampling proposal distribution is formed using the
  an ensemble of particles. A stratified sample is taken from this
  distribution and weighted under the posterior, a state-of-the-art
  resampling method is then used to create an evenly weighted sample
  ready for the next iteration. We demonstrate that this parallel
  adaptive importance sampling (PAIS) method outperforms naive
  parallelisation of serial MCMC methods for low dimensional problems,
  and in fact shows
  better than linear improvements in convergence rates with respect to
  the number of ensemble members. We also introduce a new resampling
  strategy, approximate multinomial resampling (AMR), which while not
  as accurate as other schemes is substantially less costly for large
  ensemble sizes, which can then be used in conjunction with PAIS for
  complex problems. In particular, we demonstrate this methodology's
  superior sampling for multimodal problems, such as those arising
  from inference for mixture models, and for problems with expensive
  likelihoods requiring the solution of a differential equation.
\end{abstract}
\begin{keywords}MCMC, parallel, importance sampling, Bayesian, inverse problems.
\end{keywords}
\section{Introduction}
Having first been developed in the early 1970s\cite{hastings1970monte}, Markov chain Monte Carlo (MCMC) methods have been of increasing
importance and interest in the last 20 years or so. They allow us to
sample from complex probability distributions which we would not be
able to sample from directly. In particular, these methods have
revolutionised the way in which inverse problems can be tackled,
allowing full posterior sampling when using a Bayesian framework.

@@

This problem has been tackled in a variety of different ways. One
approach is to construct increasingly complex MCMC methods which are
able to use the structure of the posterior to make more intelligent
proposals, leading to more thorough exploration of the posterior with
fewer iterations. For example, the Hamiltonian or Hybrid Monte Carlo
(HMC) algorithm uses gradient information and symplectic integrators
in order to make very large moves in state with relatively high
acceptance probability\cite{sexton1992hamiltonian}. Non-reversible
methods are also becoming quite popular as they can improve
mixing\cite{bierkens2015non}. Riemann
manifold Monte Carlo methods exploit the Riemann geometry of the
parameter space, and are able to take advantage of the local structure
of the target density to produce more efficient MCMC
proposals\cite{girolami2011riemann}. This methodology has been
successfully applied to MALA-type proposals and methods which exploit
even higher order gradient information\cite{bui2014solving}.

@@

However, high dimensionality is not the only challenge that we may
face. Complex structure in low
dimensions can cause significant issues. These issues may arise due to
large correlations between parameters in the posterior, leading to
long thin curved structures which many standard methods can struggle
with. These features are common, for example, in inverse problems
related to epidemiology and
other biological applications\cite{house2016bayesian}. Multimodality of the posterior can also lead to
incredibly slow convergence in many methods. Many methods allow for
good exploration of the current mode, but the waiting time to the next
switch of the chain to another mode may be large. Since many switches
are required in order for the correct weighting to be given to each
mode, and for all of the modes to be explored fully, this presents a
significant challenge.

One application where this is an ever-present problem is that of
mixture models. Given a dataset, where we know that the data is from
two or more different distributions, we wish to be able to identify
the parameters, e.g. the mean and variance and relative weight, of
each part of the mixture\cite{marin2005bayesian}. The resulting posterior
distribution is invariably a multimodal distribution, since the likelihood is
invariant to permutations. Metropolis-Hastings algorithms, for
example, will often fail to converge in a reasonable time frame for
problems such as this. Since the posterior may be multimodal,
independent of this label switching, it is important to be able
to efficiently sample from the whole posterior.

@@

The method considered is similar to population Monte Carlo methods
which have been previously studied, for example in
\cite{elvira2017improving}. However our approach exploits the current
state-of-the-art resamplers. Higher accuracy resamplers, that provide
a deterministic resample which optimally reflects the statistics of
the original sample, lead to better mixture approximations of the
posterior. These approximations can then be used to construct stable
and efficient importance sampling proposals. The cost of the
state-of-art resamplers can be prohibitive for large ensemble sizes,
which are necessary for sampling in moderately higher dimensions, and
as such we also present a greedy approximation of the optimal
transport resampler, which provides us with good results for a
fraction of the cost. We also detail how scaling parameters in the
MCMC proposals can be quickly tuned, leading us to an efficient and
fully-automated algorithm.

@@

The quantity $x$ is mapped on to observable space by the observation
operator $\mathcal{G}:X \to\mathbb{R}^d$. We assume that the
observations, $\data$, are subject to Gaussian noise,
\begin{equation}\label{eqn:obs}
	\data = \mathcal{G}(x) + \varepsilon, \qquad \varepsilon \sim \mu_{\varepsilon} = \mathcal{N}(0,\Sigma).
\end{equation}
For example, if $x$ is an initial condition for the wave equation,
$\mathcal{G}$ might calculate the height of a wave at a particular point and time.

These modelling assumptions allow us to construct the 
likelihood of observing the data $\data$ given the quantity $x =
x^*$. Rearranging \eqref{eqn:obs} and using the distribution of
$\varepsilon$, we get:
\begin{equation}\label{eqn:like}
	\mathbb{P}(\data|x=x^*) \propto \exp \left ( -\frac{1}{2} \|\mathcal{G}(x^*)
	  - \data\|_\Sigma^2 \right ) = \exp\left(-\Phi(x^*)\right),
\end{equation}
where $\| y_1 - y_2 \|_\Sigma = (y_1-y_2)^\top\Sigma^{-1}(y_1-y_2)$
for $y_1,y_2 \in \mathbb{R}^d$.

@@

In this work we use these particle filters as resampling methods. We
approximate the posterior distribution, $\mu_Y$, with a small sample
of weighted particles, $\{(w_i,y_i)\}_{i=1}^M$. In filtering problems
the weights would be found by incorporating new observed data whereas
here we simply use importance weights. This distribution, $\hat{\mu}_Y
\approx \mu_Y$, can then be resampled, using the ETPF or otherwise,
into a new distribution, $\hat{\eta} \approx \mu_Y$,
\begin{equation}\label{eqn:resampler}
	\hat{\mu}_Y = \sum\limits_{i=1}^M \! w_i\delta_{y_i}(\cdot)
	 \quad \xrightarrow{\text{ETPF}} \quad \sum\limits_{i=1}^M \!
	 \frac{1}{M} \delta_{x_i}(\cdot) = \hat{\eta}.
\end{equation}
where $n_i \in \{0, 1, \dots,M\}$. This
new sample $\{x_i\}_{i=1}^M$ is equally weighted.

Particle filters such as the ETPF are well suited to this problem
since it is easy to introduce conditions in the resampling to ensure
you obtain the behaviour you require. One downside is that the
required ensemble size increases quickly with dimension, making it
difficult to use in high-dimensional problems.

@@

\subsection{Deficiencies of Metropolis-type MCMC schemes}
All MCMC methods are naively parallelisable. One can take a method
and simply implement it simultaneously over an ensemble of processors. All
of the states of all of the ensemble member can be recorded, and in the time
that it takes one MCMC chain to draw $N$ samples, $M$ ensemble members
can draw $NM$ samples.

@@

Moreover, many MCMC algorithms suffer from poor mixing, especially in
multimodal systems. The amount of samples that it takes for an MCMC
trajectory to switch between modes can be large, and given that a large
number of switches are required before we have a good idea of the relative
probability densities of these different regions, it can be prohibitively
expensive.

@@

Moreover, a naively parallelised MCMC scheme is exactly that -
naive. Intuition suggests that we can gain some speed up by sharing
information across the ensemble members, and this is
what we wish to demonstrate in this paper.

@@

The proposal distribution could be constructed in many different ways,
but we choose to use a mixture distribution, made up of a sum of MCMC
proposal kernels. This corresponds to replacing the Dirac distribution
in the right-hand side of Equation~\ref{eqn:resampler} with a chosen
distribution e.g. Gaussian. Once the proposal is constructed, we can
sample a new ensemble from the proposal distribution, and each is
assigned a weight given by the ratio of the target density and the
proposal mixture density. Assuming that our proposal distribution is a
good one, then the variance of the weights will be small, and we will
have many useful samples. Finally, we need to create a set of evenly
weighted samples which best represent this set of weighted samples.
This is achieved by implementing a resampling algorithm. These samples
are not stored in order to characterise the posterior density, since
the resampling process inevitably adds some error/bias. They are
simply needed in order to inform the mixture distribution for the next
iteration of the algorithm.

Initially we
will use the ETPF algorithm\cite{reich2013nonparametric}, although we
will suggest an alternative strategy in Section \ref{sec:AMR}. The
resampling algorithm gives us a set of evenly weighted samples which
represents the target distribution well, which we can use to iterate
the process again. The algorithm is summarised in Algorithm
\ref{alg:PAIS}.

@@

Since the resampling does not give us a statistically identical sample
to that which is inputted, we cannot assume that the samples $\X^{(i)}$
are samples from the posterior. Therefore, as with serial
importance samplers, the weighted samples
$(\W^{(i)},\Y^{(i)})_{i=1}^N$ are the samples from the posterior that
we will analyse.

@@

The key is to choose a suitable transition kernel $\nu$ such that
if $X^{(i)}$ is a representative sample of the posterior,
then the mixture density $\chi(\cdot;\X^{(i)})$ is a good
approximation of the posterior distribution. If this is the case,
the newly proposed states $\Y^{(i)}$ will also be a good sample of the
posterior with low variance in the weights $\W^{(i)}$.

In Section \ref{Sec:Num}, we will demonstrate how the algorithm
performs, primarily using random walk (RW) proposals. We do not claim that this choice is optimal, but
is simply chosen as an example to show that sharing information across
ensemble members can improve on the original MH algorithm and lead to
tolerances being achieved in fewer evaluations of $\G$. This is important since if
the inverse problem being tackled involves computing
the likelihood from a very large data set this could lead to a
large saving of computational cost. We have observed that
using more complex (and hence more expensive) kernels $\nu$, does not
significantly improve the speed of convergence of the algorithm for
posteriors that we have considered\cite{Paul}.

Care needs to be taken when choosing the proposal distribution to
ensure that the proposals are absolutely continuous with respect to
the posterior distribution. If this is not the case, importance
weights grow exponentially for samples in the tails. This will
drastically decrease the efficiency of sampling.

@@

Most MCMC proposals have parametric dependence which
allows the user to control their variance. For example, in the RW
proposal $y = x + \beta \eta$, the parameter $\beta$ is the standard deviation
of the proposal distribution. Therefore the proposal distributions can
be tuned such that they are are slightly over-dispersed. This tuning
can take place during the burn-in phase of the algorithm. Algorithms
which use this method to find optimal proposal distributions are known
as adaptive MCMC algorithms, and have been shown to be convergent
provided that they satisfy certain
conditions\cite{roberts2007coupling,roberts2009examples}.
Alternatively, a costly trial and error scheme with short runs of the
MCMC algorithm can be used to find an acceptable value of $\beta$.

@@

The effective sample size, $\neff$, can be used to assess the
efficiency of importance samplers. Ideally, in each iteration, we
would like all $M$ of our samples to provide us with new information
about the posterior distribution. In practise, we cannot achieve a
perfect effective sample size of $M$.

@@

The statistic $\neff$ is easier to deal with than the variance of the
weights, as it takes values on $[1, M]$ while the variance of the
weights takes values on $\mathbb{R}_\geq$. Moreover the variance of
the weights can vary
over many orders of magnitude causing numerical issues, so that the effective
sample size is more desirable as an indicator of optimality.

@@

Although the ETPF is optimal in terms of preserving statistics of the
sample, it can also become quite costly as the number of ensemble members
is increased. It is arguable that in the context of PAIS, we do not
require this degree of accuracy, and that a faster more approximate
method for resampling could be employed. One approach would be to use
the bootstrap resampler, which simply takes the $M$ ensemble members'
weights and constructs a multinomial distribution, from which $M$
samples are drawn. This is essentially the cheapest resampling
algorithm that one could construct. However it too has some
drawbacks. The algorithm is random, and as such it is possible for all
of the ensemble members in a particular region not to be sampled. This
could be particularly problematic when attempting to sample from a
multimodal distribution, where it might take a long time to find one
of the modes again. The bootstrap filter is also not guaranteed to
preserve the mean of the weighted sample, unlike the ETPF.

Ideally, we would like to use a resampling algorithm which is not
prohibitively costly for moderately or large sized ensembles,
which preserves the mean of the samples, and which makes it much
harder for the new samples to forget a significant region in the
density. This motivates the following algorithm, which we refer to as
approximate multinomial resampling (AMR).

@@

To test the accuracy and speed of the three resamplers (ETPF,
bootstrap and AMR), we drew a sample of size $M$ from the proposal distribution
$\mathcal{N}(1,2)$. Importance weights were assigned, based on a
target distribution of $\mathcal{N}(2,3)$. The statistics of the
resampled outputs were compared with the original weighted samples. Figure \ref{fig:AMR} (a)-(c) show how the relative errors in the first
three moments of the samples changes with ensemble size $M$ for the three different
samplers. As expected, the AMR lies somewhere between the high
accuracy of the ETPF and the less accurate bootstrap
resampling. Note that only the error for the bootstrap multinomial
sampler is presented for the first moment since both the ETPF and the
AMR preserve the mean of the original weighted samples up to machine precision. Figure \ref{fig:AMR} (d) shows how the computational cost,
measured in seconds, scales with the ensemble size for the three
different methods, where timings have been taken from simulations on a Dell
server with four 8-core 3.3GHz CPUs and 64Gb memory. These results demonstrate that the AMR behaves how we wish, and
importantly ensures that exactly one sample of the output will lie in
each region with weights up to $\frac{1}{M}$ of the total.

We will use the AMR in the numerics in Section \ref{sec:mixture}
where we have chosen to use a larger ensemble size. We do not claim
that the AMR is the optimal choice within PAIS, but it does have
favourable features, and demonstrates how different choices of
resampler can affect the speed and accuracy of the PAIS algorithm.

@@

\section{Numerical Examples}\label{Sec:Num}
The numerical examples in this section were all computed on a single
core of Dell
server with four 8-core 3.3GHz CPUs and 64Gb memory. Throughout we
measure the efficiency of the algorithms in terms of the number of
iterations (i.e. likelihood evaluations) which are required in order
to reach a given order of accuracy. The reason for this is that this
method is designed with a particular type of challenging inverse
problem in mind, namely one which is low dimensional, has an expensive
likelihood (such as a PDE solve, and/or very large data) which dwarfs
the overhead cost of PAIS, and which could have complex posterior
structure, such as correlations, multimodality or other highly
non-Gaussian features.

@@

\begin{figure}[ht]
\begin{center}
\includegraphics[width=0.6\textwidth]{"figures/PAIS_saving"}
\caption{Ratio of PAIS-RW samples required to reach the same tolerance as the RWMH algorithm.}
\label{fig:PAIS_saving}
\end{center}
\end{figure}

@@

In this example we look at a bimodal posterior distribution which has
a high energy barrier between the two modes. In the MH algorithm, the
expected number of iterations for a switch between modes is a large
proportion of the total number of iterations. This posterior is presented in
Figure~\ref{fig:BM2_posterior}.

@@

\begin{table}[!htb]
      \centering
        \begin{tabular}{|l|r|r|r|}
	\hline
	Algorithm							& $\beta^*_{\text{acc}}$	& $\beta^*_{\text{eff}}$	& $\beta^*_{\text{L2}}$ \\ \hline
	RWMH								& 2.3e-1					& - 						& 9.3e-1\\
	PAIS-RW								& -						& 5.1e-2 					& 1.3e-1\\
	\hline
	\end{tabular}
	\vspace{2mm}
	\caption{Optimal scaling parameters for the example in Section~\ref{sec:bimodal}.}
	\label{tab:BM2_opt_delta}
\end{table}

@@

The MH and PAIS algorithms are run with both the globally and locally
optimal scaling parameters. Figure~\ref{fig:BM2_L2} (a) shows that the
globally optimal $\beta^*$ leads to normal convergence rates,
$\mathcal{O}(n^{-1/2})$, whereas algorithms using locally optimal
scaling parameters initially converge faster but do not sample
proportionally for the whole simulation.

Using the PAIS algorithm we can ensure that we do not lose modes as
simulation progresses. Since we have an ensemble, we can sacrifice
locally optimal sampling for one or two ensemble members in favour of
a larger proposal variance. These chains with larger scaling
parameters act both as `scouts' for new modes, and act to aid in the
over-dispersal of the proposal distribution. Figure~\ref{fig:BM2_L2}
(b) shows the results of using 49 chains with the local optimal
scaling parameter, and one chain with ten times the local optimal
scaling parameter. We see that modes are not forgotten and the
algorithm converges with the improvement we see from PAIS-RW in the
Gaussian example.

@@

The technique of mixture modelling employs well known parametric
families to construct an approximating nonparametric distribution
which may have a complex structure. Most commonly, Gaussian kernels
are used since underlying properties in the data can often be assumed
to follow a Gaussian distribution. An example would be if a
practitioner were to measure the heights of one hundred adults, but
failed to record their gender. The data could be considered as one
population with two sub populations, male and female. The problem then
might be to find the average height of adult males from the data. In
this case, since height is often considered to follow a Gaussian
distribution, it makes sense to model the population as a mixture of
two univariate Gaussian distributions.

@@

Figure \ref{fig:mixture_posterior} presents a visualisation of the
posterior distribution for this problem, created from 10million
samples produced by the PAIS-RW algorithm.

>>>>

\author{Colin Cotter\thanks{Department of Mathematics, Imperial
    College, London, UK} \and Simon Cotter\thanks{School of
    Mathematics, University of Manchester, Manchester, UK. e:
    simon.cotter@manchester.ac.uk. SLC is grateful for EPSRC First
    grant award EP/L023393/1} \and Paul Russell\thanks{School of
    Mathematics, University of Manchester, Manchester, UK}}
\title{Parallel Adaptive Importance Sampling}
\begin{document}
\maketitle
\begin{abstract}
  Markov chain Monte Carlo methods are a powerful and commonly used
  family of numerical methods for sampling from complex probability
  distributions. As applications of these methods increase in size and
  complexity, the need for efficient methods which can exploit the
  parallel architectures which are prevalent in high performance
  computing increases. In this paper, we aim to develop a framework
  for scalable parallel sampling algorithms. At each iteration, an
  importance sampling proposal distribution is formed using the
  ensemble of particles. A stratified sample is taken from this
  distribution and weighted under the posterior, a state-of-the-art
  resampling method is then used to create an evenly weighted sample
  ready for the next iteration. We demonstrate that this parallel
  adaptive importance sampling (PAIS) method outperforms naive
  parallelisation of serial MCMC methods for low dimensional problems,
  and in fact shows
  better than linear improvements in convergence rates with respect to
  the number of ensemble members. We also introduce a new resampling
  strategy, approximate multinomial resampling (AMR), which while not
  as accurate as other schemes is substantially less costly for large
  ensemble sizes, which can then be used in conjunction with PAIS for
  complex problems. We also focus on how algorithmic parameters
  regarding the mixture proposal can be quickly tuned to optimise
  performance. In particular, we demonstrate this methodology's
  superior sampling for multimodal problems, such as those arising
  from inference for mixture models, and for problems with expensive
  likelihoods requiring the solution of a differential equation.
\end{abstract}
\begin{keywords}MCMC, parallel, importance sampling, Bayesian, inverse problems.
\end{keywords}
\section{Introduction}
Having first been developed in the early 1970s\cite{hastings1970monte}, Markov chain Monte Carlo (MCMC) methods have been of increasing
importance and interest in the last 20 years or so. They allow us to
sample from complex probability distributions which we would not be
able to sample from directly. In particular, these methods have
revolutionised the way in which inverse problems can be tackled,
allowing full posterior sampling when using a Bayesian framework.

@@

This problem has been tackled in a variety of different ways. One
approach is to construct increasingly complex MCMC methods which are
able to use the structure of the posterior to make more intelligent
proposals, leading to more thorough exploration of the posterior with
fewer iterations. For example, the Hamiltonian or Hybrid Monte Carlo
(HMC) algorithm uses gradient information and symplectic integrators
in order to make very large moves in state with relatively high
acceptance probabilities\cite{sexton1992hamiltonian}. Non-reversible
methods are also becoming quite popular as they can improve
mixing\cite{bierkens2015non}. Riemann
manifold Monte Carlo methods exploit the Riemann geometry of the
parameter space, and are able to take advantage of the local structure
of the target density to produce more efficient MCMC
proposals\cite{girolami2011riemann}. This methodology has been
successfully applied to MALA-type proposals and methods which exploit
even higher order gradient information\cite{bui2014solving}.

@@

However, high dimensionality is not the only challenge that we may
face. Complex structure in low
dimensions can cause significant issues. These issues may arise due to
large correlations between parameters in the posterior, leading to
long thin curved structures which many standard methods can struggle
with. These features are common, for example, in inverse problems
related to epidemiology and
other biological applications\cite{house2016bayesian}.

Multimodality of the posterior can also lead to
incredibly slow convergence in many methods. Many methods allow for
good exploration of the current mode, but the waiting time to the next
switch of the chain to another mode may be large. Since many switches
are required in order for the correct weighting to be given to each
mode, and for all of the modes to be explored fully, this presents a
significant challenge. One application where this is an ever-present problem is that of
mixture models. Given a dataset, where we know that the data is from
two or more different distributions, we wish to be able to identify
the parameters, e.g. the mean and variance and relative weight, of
each part of the mixture\cite{marin2005bayesian}. The resulting posterior
distribution is invariably a multimodal distribution, since the likelihood is
invariant to permutations. Metropolis-Hastings algorithms, for
example, will often fail to converge in a reasonable time frame for
problems such as this. Since the posterior may be multimodal,
independent of this label switching, it is important to be able
to efficiently sample from the whole posterior.

@@

The method considered is similar to population Monte Carlo methods
which have been previously studied, for example in
\cite{elvira2017improving}. However our approach exploits the current
state-of-the-art resamplers. Higher accuracy resamplers, that provide
a deterministic resample which optimally reflects the statistics of
the original sample, lead to better mixture approximations of the
posterior. These approximations can then be used to construct stable
and efficient importance sampling proposals. The cost of the
state-of-art resamplers can be prohibitive for large ensemble sizes,
which are necessary for sampling in moderately higher dimensions, and
as such we also present a greedy approximation of the optimal
transport resampler, which provides us with good results for a
fraction of the cost. We also detail how scaling parameters in the
MCMC proposals can be quickly tuned, leading us to an efficient and
fully-automated algorithm. We will demonstrate that despite additional
overheads, the PAIS methodology can outperform its plain
Metropolis-Hastings cousin by orders of magnitude to reach a histogram
of a particular degree of convergence.

@@

The quantity $x$ is mapped on to observable space by the observation
operator $\mathcal{G}:X \to\mathbb{R}^d$. We assume that the
observations, $\data$, are subject to Gaussian noise,
\begin{equation}\label{eqn:obs}
	\data = \mathcal{G}(x) + \varepsilon, \qquad \varepsilon \sim \mu_{\varepsilon} = \mathcal{N}(0,\Sigma).
\end{equation}

These modelling assumptions allow us to construct the 
likelihood of observing the data $\data$ given the quantity $x =
x^*$. Rearranging \eqref{eqn:obs} and using the distribution of
$\varepsilon$, we get:
\begin{equation}\label{eqn:like}
	\mathbb{P}(\data|x=x^*) \propto \exp \left ( -\frac{1}{2} \|\mathcal{G}(x^*)
	  - \data\|_\Sigma^2 \right ) = \exp\left(-\Phi(x^*)\right),
\end{equation}
where $\| y_1 - y_2 \|^2_\Sigma = (y_1-y_2)^\top\Sigma^{-1}(y_1-y_2)$
for $y_1,y_2 \in \mathbb{R}^d$.

@@

In this work we use the resampler used within these particle filters. We
approximate the posterior distribution, $\mu_Y$, with a sample
of weighted particles, $\{(w_i,y_i)\}_{i=1}^M$. In filtering problems
the weights would be found by incorporating new observed data whereas
here we simply use importance weights. This distribution, $\hat{\mu}_Y
\approx \mu_Y$, can then be resampled, using the ETPF or otherwise,
into a new distribution, $\hat{\eta} \approx \mu_Y$,
\begin{equation}\label{eqn:resampler}
	\hat{\mu}_Y = \sum\limits_{i=1}^M \! w_i\delta_{y_i}(\cdot)
	 \quad \xrightarrow{\text{ETPF}} \quad \sum\limits_{i=1}^M \!
	 \frac{1}{M} \delta_{x_i}(\cdot) = \hat{\eta}.
\end{equation}
where $n_i \in \{0, 1, \dots,M\}$. This
new sample $\{x_i\}_{i=1}^M$ is equally weighted.

The resampler used within particle filters such as the ETPF are well suited to this problem
since it is easy to introduce conditions in the resampling to ensure
you obtain the behaviour you require. One downside is that the
required ensemble size increases quickly with dimension, making it
difficult to use in high-dimensional problems.

@@

\subsection{Deficiencies of Metropolis-type MCMC schemes}
All MCMC methods are naively parallelisable. One can take a method
and simply implement it multiple times over an ensemble of processors. All
of the states of all of the ensemble member can be recorded, and in the time
that it takes one MCMC chain to draw $N$ samples, $M$ ensemble members
can draw $NM$ samples.

@@

Moreover, many MCMC algorithms suffer from poor mixing, especially in
multimodal systems. The number of samples that it takes for an MCMC
trajectory to switch between modes can be large, and given that a large
number of switches are required before we have a good idea of the relative
probability densities of these different regions, it can be prohibitively
expensive.

@@

@@

The proposal distribution could be constructed in many different ways,
but we choose to use a mixture distribution, made up of a sum of MCMC
proposal kernels  e.g. Gaussians centred at the current state, as in RWMH. Once the proposal is constructed, we can
sample a new ensemble from the proposal distribution, and each is
assigned a weight given by the ratio of the target density and the
proposal mixture density. Assuming that our proposal distribution is a
good one, then the variance of the weights will be small, and we will
have many useful samples. Finally, we need to create a set of evenly
weighted samples which best represent this set of weighted samples.
This is achieved by implementing a resampling algorithm. These samples
are not stored in order to characterise the posterior density, since
the resampling process inevitably adds some error. They are
simply needed in order to inform the mixture proposal distribution for the next
iteration of the algorithm.

Initially we
will use the ETPF resampler algorithm\cite{reich2013nonparametric}, although we
will suggest an alternative strategy in Section \ref{sec:AMR} where a
large ensemble is required, and for which the ETPF may become more expensive. The
resampling algorithm gives us a set of evenly weighted samples which
represents the target distribution well, which we can use to iterate
the process again. The algorithm is summarised in Algorithm
\ref{alg:PAIS}.

@@

Since the resampling does not give us a statistically identical sample
to that which is input, we cannot assume that the samples $\X^{(i)}$
are samples from the posterior. Therefore, as with serial
importance samplers, the weighted samples
$(\W,\Y)_{i=1}^N$ are the samples from the posterior that
we will analyse.

@@

The key is to choose a suitable transition kernel $\nu$ such that
if $\mathbf{X}^{(i)}$ is a representative sample of the posterior,
then the mixture density $\chi(\cdot;\X^{(i)})$ is a good
approximation of the posterior distribution. If this is the case,
the newly proposed states $\Y^{(i)}$ will also be a good sample of the
posterior with low variance in the weights $\W^{(i)}$.

In Section \ref{Sec:Num}, we will demonstrate how the algorithm
performs, primarily using random walk (RW) proposals. We do not claim that this choice is optimal, but
is simply chosen as an example to show that sharing information across
ensemble members can improve on the original MH algorithm and lead to
tolerances being achieved in fewer evaluations of $\G$. This is important since if
the inverse problem being tackled involves computing
the likelihood from a very large data set, or where the likelihood
requires the numerical solution of a differential equation, this could lead to a
large saving of computational cost. We have observed that
using more complex (and hence more expensive) kernels $\nu$, does not
significantly improve the speed of convergence of the algorithm for
posteriors that we have considered, although kernels such as those
used in MALA can be more stable for certain problems\cite{Paul}.

Care needs to be taken when choosing kernel(s) for the proposal distribution to
ensure that the overall mixture is absolutely continuous with respect to
the posterior distribution. If this is not the case, importance
weights can become high variance where samples are made away from
modes in the distribution. This will
drastically decrease the efficiency of sampling.

@@

Most MCMC proposals have parametric dependence which
allows the user to control their variance. For example, in the RW
proposal $y = x + \beta \eta$, the parameter $\beta$ is the standard deviation
of the proposal distribution. Therefore the proposal distributions can
be tuned such that they are slightly over-dispersed. This tuning
can take place during the burn-in phase of the algorithm. Algorithms
which use this method to find optimal proposal distributions are known
as adaptive MCMC algorithms, and have been shown to be convergent
provided that they satisfy certain
conditions\cite{roberts2007coupling,roberts2009examples}.
Alternatively, a costly trial and error scheme with short runs of the
MCMC algorithm can be used to find an acceptable value of $\beta$.

@@

The effective sample size, $\neff$, can be used to assess the
efficiency of importance samplers. Ideally, in each iteration, we
would like all $M$ of our samples to provide us with new information
about the posterior distribution. In practice, we cannot achieve a
perfect effective sample size of $M$.

@@

The statistic $\neff$ is easier to deal with than the variance of the
weights, as it takes values on $[1, M]$ while the variance of the
weights takes values on $\mathbb{R}_\geq$. Moreover the variance of
the weights can vary
over many orders of magnitude causing numerical instabilities, so that the effective
sample size is more desirable as an indicator of optimality.

@@

Although the ETPF produces the optimal linear coupling, it can also become quite costly as the number of ensemble members
is increased. It is arguable that in the context of PAIS, we do not
require this degree of accuracy, and that a faster more approximate
method for resampling could be employed. One approach would be to use
the bootstrap resampler, which simply takes the $M$ ensemble members'
weights and constructs a multinomial distribution, from which $M$
samples are drawn. This is essentially the cheapest resampling
algorithm that one could construct. However it too has some
drawbacks. The algorithm is random, and as such it is possible for all
of the ensemble members in a particular region not to be sampled. This
could be particularly problematic when attempting to sample from a
multimodal distribution, where it might take a long time to find one
of the modes again. The bootstrap filter is also not guaranteed to
preserve the mean of the weighted sample, unlike the ETPF.

Ideally, we would like to use a resampling algorithm which is not
prohibitively costly for moderately or large sized ensembles,
which preserves the mean of the samples, and which makes it much
harder for the new samples to forget a significant region in the
density. This motivates the following algorithm, which we refer to as the
approximate multinomial resampler (AMR), which is a greedy
approximation of the ETPF resampler.

@@

To test the accuracy and speed of the three resamplers (ETPF,
bootstrap and AMR), we drew a sample of size $M$ from the proposal distribution
$\mathcal{N}(1,2)$. Importance weights were assigned, based on a
target distribution of $\mathcal{N}(2,3)$. The statistics of the
resampled outputs were compared with the original weighted samples. Figure \ref{fig:AMR} (a)-(c) show how the relative errors in the first
three moments of the samples changes with ensemble size $M$ for the three different
resamplers. As expected, the AMR lies somewhere between the high
accuracy of the ETPF and the less accurate bootstrap
resampling. Note that only the error for the bootstrap multinomial
sampler is presented for the first moment since both the ETPF and the
AMR preserve the mean of the original weighted samples up to machine precision. Figure \ref{fig:AMR} (d) shows how the computational cost,
measured in seconds, scales with the ensemble size for the three
different methods, where timings have been taken from simulations on a Dell
server with four 8-core 3.3GHz CPUs and 64Gb memory. These results demonstrate that the AMR behaves how we wish, and
importantly ensures that exactly one sample of the output will lie in
each region with weights up to $\frac{1}{M}$ of the total.

We will use the AMR in the numerics in Section \ref{sec:mixture}
where we have chosen to use a larger ensemble size. We do not claim
that the AMR is the optimal choice within PAIS, but it does have
favourable features, and demonstrates how different choices of
resampler can affect the speed and efficiency of the PAIS algorithm.

@@

\section{Numerical Examples}\label{Sec:Num}
The numerical examples in this section were all computed on a single
core of a Dell
server with four 8-core 3.3GHz CPUs and 64Gb memory. Throughout we
measure the efficiency of the algorithms in terms of the number of
iterations (i.e. likelihood evaluations) which are required in order
to reach a given order of accuracy. The reason for this is that this
method is designed with a particular type of challenging inverse
problem in mind, namely one which is low dimensional, has an expensive
likelihood (such as a PDE solve, and/or very large data) which dwarfs
the overhead cost of PAIS, and which could have complex posterior
structure, such as correlations, multimodality or other highly
non-Gaussian features.

@@

\begin{figure}[ht]
\begin{center}
\includegraphics[width=0.6\textwidth]{"figures/PAIS_saving"}
\caption{Ratio of PAIS-RW samples required to reach the same tolerance
  as the RWMH algorithm, using the ETPF resampler.}
\label{fig:PAIS_saving}
\end{center}
\end{figure}

@@

In this example we look at a bimodal posterior distribution which has
a region of very low probability density between the two modes. In the MH algorithm, the
expected number of iterations for a switch between modes is a large
proportion of the total number of iterations. This posterior is presented in
Figure~\ref{fig:BM2_posterior}.

@@

\begin{table}[!htb]
      \centering
        \begin{tabular}{|l|r|r|r|}
	\hline
	Algorithm							& $\beta^*_{\text{acc}}$	& $\beta^*_{\text{eff}}$	& $\beta^*_{\text{L}^2}$ \\ \hline
	RWMH								& 2.3e-1					& - 						& 9.3e-1\\
	PAIS-RW								& -						& 5.1e-2 					& 1.3e-1\\
	\hline
	\end{tabular}
	\vspace{2mm}
	\caption{Optimal scaling parameters for the example in Section~\ref{sec:bimodal}.}
	\label{tab:BM2_opt_delta}
\end{table}

@@

The MH and PAIS algorithms are run with both the globally and locally
optimal scaling parameters. Figure~\ref{fig:BM2_L2} (a) shows that the
globally optimal $\beta^*$ leads to normal convergence rates,
$\mathcal{O}(1/\sqrt{N})$, whereas algorithms using locally optimal
scaling parameters do not sample
proportionally for the whole simulation.

Using the PAIS algorithm we can ensure that we do not lose modes as the
simulation progresses. Since we have an ensemble, we can sacrifice
locally optimal sampling for one or two ensemble members in favour of
a larger proposal variance. These chains with larger scaling
parameters act both as `scouts' for new modes, and act to aid in the
over-dispersal of the proposal distribution. Figure~\ref{fig:BM2_L2}
(b) shows the results of using 49 chains with the local optimal
scaling parameter, and one chain with ten times the local optimal
scaling parameter. We see that modes are not forgotten and the
algorithm converges with the improvement we see from PAIS-RW in the
Gaussian example.

@@

The technique of mixture modelling employs well known parametric
families to construct an approximating distribution
which may have a complex structure. Most commonly, Gaussian kernels
are used since underlying properties in the data can often be assumed
to follow a Gaussian distribution. An example would be if a
practitioner were to measure the heights of one hundred adults, but
failed to record their gender. The data could be considered as one
population with two sub populations, male and female. The problem then
might be to find the average height of adult males from the data. In
this case, since height is often considered to follow a Gaussian
distribution, it makes sense to model the population as a mixture of
two univariate Gaussian distributions.

@@

Figure \ref{fig:mixture_posterior} presents a visualisation of the
posterior distribution for this problem, created from 10 million
samples produced by the PAIS-RW algorithm.
